---
title: "DNase Dataset Exploration using Mixed Models"
output: html_notebook
---

# Data Exploration

The DNase dataset in R is a collection of data in the development of an assay.
According to the R documentation,

An object of class c("nfnGroupedData", "nfGroupedData", "groupedData", "data.frame") containing the following columns:

Run      - an ordered factor with levels 10 < … < 3 indicating the assay run.
conc     - a numeric vector giving the known concentration of the protein.
density  - a numeric vector giving the measured optical density (dimensionless)
           in the assay. Duplicate optical density measurements were obtained.


The description for the factors `Run` is not exactly correct as the ordered
factors includes 11
```{r}
unique(DNase$Run)
```

What does the data look like "raw"

```{r}
plot(DNase)
```
What we see is that there are duplicates in the density row. What is their structure?
```{r}
head(DNase, 10)
```
The duplicate density measurement occur in pairs of identical protein 
concentration. Let's look at the density as a function of concentration.

```{r}
plot(
    x=DNase$conc,
    y=DNase$density,
    xlab="Concentration",
    ylab="Density"
)
```

The trend indicates there is saturation limit, unless the true growth is
logarithmic. For modeling purposes, let's consider logistic growth. What does a
`tanh` model look like?

```{r}
tanh_input_seq <- seq(0, 12, length.out=50)
plot(tanh_input_seq, 0.01 + 2.0 * tanh(0.2 * tanh_input_seq))
```

I would characterize this model as sufficient given some optimization.
Given the density measurements are in pairs, let's split the data frame into
training and testing sets.

```{r}
train_seq <- seq(1, nrow(DNase)-1, by=2)
test_seq <- seq(2, nrow(DNase), by=2)
if (nrow(DNase) %% 2 == 1){
    train_seq <- c(train_seq, nrow(DNase))
}

DNase_train <- DNase[train_seq,]
DNase_test <- DNase[test_seq, ]
```

Let's now try to fit a logistic model to the entire training dataset naively.

$Y = \beta_1 \tanh{\left(\beta_2 X + \beta_3\right)}$

where $X$ is the concentration and $Y$ is the density.

```{r}
density_model <- function(X, b1, b2, b3){
    return(b1 * tanh(b2 * X + b3))
}

model_fit <- nls(
    density ~ density_model(conc, b1, b2, b3),
    data = DNase_train,
    start = list(b1=2., b2=0.2, b3=0.01)
)

summary(model_fit)
```

```{r}
model_predictions <- predict(
    model_fit,
    DNase_train$conc
)
plot(DNase_test$conc, DNase_test$density)
lines(DNase_train$conc, model_predictions)
```
There are some sloppy bookkeeping as the first and last line points intersect,
but that is OK as we want to get a feel for the model.

What does each level fit have to tell us? Does each `Run` have a similar
fit quality?

```{r}
plot(DNase_train$conc, DNase_train$density, col="black", xlim=range(DNase$conc), ylim=range(DNase$density))
plot(DNase_test$conc, DNase_test$density, col="black", xlim=range(DNase$conc), ylim=range(DNase$density))
for(run_level in unique(DNase_train$Run)){
    train_data_for_model <- DNase_train[DNase_train$Run == run_level, ]
    test_data_for_model <- DNase_test[DNase_test$Run == run_level, ]
    temp_model_fit <- nls(
        density ~ density_model(conc, b1, b2, b3),
        data = train_data_for_model,
        start = list(b1=2., b2=0.2, b3=0.01)
    )
    temp_model_predictions <- predict(
        temp_model_fit,
        test_data_for_model$conc
    )
    lines(test_data_for_model$conc, temp_model_predictions)
}
```

# Mixed Models
