---
title: "DNase Dataset Exploration using Mixed Models"
output: html_notebook
---

# Data Exploration

The DNase dataset in R is a collection of data in the development of an assay.
According to the R documentation,

An object of class c("nfnGroupedData", "nfGroupedData", "groupedData", "data.frame") containing the following columns:

Run      - an ordered factor with levels 10 < … < 3 indicating the assay run.
conc     - a numeric vector giving the known concentration of the protein.
density  - a numeric vector giving the measured optical density (dimensionless)
           in the assay. Duplicate optical density measurements were obtained.


The description for the factors `Run` is not exactly correct as the ordered
factors includes 11
```{r}
unique(DNase$Run)
```

What does the data look like "raw"

```{r}
plot(DNase)
```
What we see is that there are duplicates in the density row. What is their structure?
```{r}
head(DNase, 10)
```
The duplicate density measurement occur in pairs of identical protein 
concentration. Let's look at the density as a function of concentration.

```{r}
plot(
    x=DNase$conc,
    y=DNase$density,
    xlab="Concentration",
    ylab="Density"
)
```

The trend indicates there is saturation limit, unless the true growth is
logarithmic. For modeling purposes, let's consider logistic growth. What does a
logistic model look like? For this analysis, let's use the self start model as we will want the gradient calculation also.

```{r}
input_seq <- seq(0, 12, length.out=50)

initial_pars <- c(
    Asym = 2.0,
    xmid = 2.0,
    scal = 1.0
)

plot(
    input_seq,
    SSlogis(
        input_seq,
        Asym=initial_pars[["Asym"]],
        xmid=initial_pars[["xmid"]],
        scal=initial_pars[["scal"]]
    ),
    xlab="Concentration",
    ylab="SSlogis ~ density"
)
```

I would characterize this model as sufficient given some optimization.
Given the density measurements are in pairs, let's split the data frame into
training and testing sets.

```{r}
train_seq <- seq(1, nrow(DNase)-1, by=2)
test_seq <- seq(2, nrow(DNase), by=2)

if (nrow(DNase) %% 2 == 1){
    train_seq <- c(train_seq, nrow(DNase))
}

DNase_train <- DNase[train_seq,]
DNase_test <- DNase[test_seq, ]
```

Let's now try to fit a logistic model `SSlogis` to the entire training dataset
naively.

$Y = \frac{\text{A}}{1+\exp{\left[\phi^{-1}\left(x_0-X\right)\right]}}$

where $Y$ is the density, $A$ is the asymptotic value, $x_0$ is the inflection point, $\phi$ is the scale, and $X$ is the concentration.

```{r}

model_fit <- nls(
    density ~ SSlogis(conc, Asym, xmid, scal),
    data = DNase_train,
    start = initial_pars
)

summary(model_fit)
```

```{r}
model_predictions <- predict(
    model_fit,
    DNase_train$conc
)
plot(DNase_test$conc, DNase_test$density)
lines(DNase_train$conc, model_predictions)
```
There are some sloppy bookkeeping as the first and last line points intersect,
but that is OK as we want to get a feel for the model.

What does each level fit have to tell us? Does each `Run` have a similar
fit quality?

```{r}
plot(DNase_train$conc, DNase_train$density, col="black", xlim=range(DNase$conc), ylim=range(DNase$density))
plot(DNase_test$conc, DNase_test$density, col="black", xlim=range(DNase$conc), ylim=range(DNase$density))
for(run_level in unique(DNase_train$Run)){
    train_data_for_model <- DNase_train[DNase_train$Run == run_level, ]
    test_data_for_model <- DNase_test[DNase_test$Run == run_level, ]
    temp_model_fit <- nls(
        density ~ SSlogis(conc, Asym, xmid, scal),
        data = train_data_for_model,
        start = initial_pars
    )
    temp_model_predictions <- predict(
        temp_model_fit,
        test_data_for_model$conc
    )
    lines(test_data_for_model$conc, temp_model_predictions)
}
```

# Mixed Models

Let's try to fit a nonlinear mixed model with a single random effect over the
training set. The library that I will use is the `lme4`.

```{r}
library(lme4)
```

I will include the same non-linear model with a random additive magnitude $u$

$Y = \frac{\text{A+u}}{1+\exp{\left[\phi^{-1}\left(x_0-X\right)\right]}}$

which is a random effects vector where $u \sim \mathcal{N}\left(0, \Sigma\right)$

```{r}
DNase_train$Run <- as.numeric(DNase_train$Run)

mixed_model <- nlmer(
    density ~ SSlogis(conc, Asym, xmid, scal) ~ (Asym|Run),
    data = DNase_train,
    start = initial_pars
)

nm1a <- update(mixed_model, nAGQ = 0L)
```

```{r}
summary(nm1a)
```
Let's examine the fixed-effects parameters against the test set
```{r}
nm1a_fexef <- fixef(nm1a)
conc_range <- range(DNase$conc)
plot_seq <- seq(conc_range[1], conc_range[2], length.out = 50)
plot(DNase_test$conc, DNase_test$density)

lines(
    plot_seq,
    SSlogis(
        plot_seq,
        nm1a_fexef[["Asym"]],
        nm1a_fexef[["xmid"]],
        nm1a_fexef[["scal"]]
    )
)
```

So how did the assay `Run` samples, the random effect, impact the fit?

```{r}
raef_out <- ranef(nm1a)$Run

raef_df <- data.frame(
    mean=raef_out[, "Asym"],
    std=rep(sigma(nm1a), nrow(raef_out))
    
)
raef_df
```


```{r}
library(ggplot2)

raef_plot <- ggplot(
    raef_df,
    aes(
        x=c(1:nrow(raef_df)),
        y=mean,
        ymin=mean-std,
        ymax=mean+std
    )
)
    # + geom_line()
    # + geom_point()
    # + geom_errorbar(aes(ymin=mean-std, ymax=mean+std), width=.2,
                 # position=position_dodge(0.05))

# print(raef_plot)
raef_plot + geom_pointrange()
```

```{r}
colMeans(raef_df)
```

As expected, the expectation of the random effect vector $E[u] = 0$.

Turns out that the standard deviation of the random effect is much larger than
the effect itself. Essentially, we see that the `Run` random effect is minimal
using this model.